A Framework for Substructural Type Systems \thanks{Supported by organization x.}

### 2021-10-13 tex/esop22/algebra.tex
[ writing ] more algebra (modules, maps)

We also make use, particularly in the mechanisation, of an alternative
definition based on relations.

\begin{definition}
  A \emph{relational posemiring} is a tuple
  $(\Ann, \leq, {{-}\leq}0, {{-}\leq}[{-}+{-}],
  {{-}\leq}1, {{-}\leq}[{-}*{-}])$
  such that the following laws hold (with free variables universally
  quantified).
  \begin{itemize}
    \item $x' \leq x \to {x \leq}0 \to {x' \leq}0$
    \item $x' \leq x \to y \leq y' \to z \leq z' \to
      {x \leq}[y+z] \to {x' \leq}[y'+z']$
    \item $x' \leq x \to {x \leq}1 \to {x' \leq}1$
    \item $x' \leq x \to y \leq y' \to z \leq z' \to
      {x \leq}[y*z] \to {x' \leq}[y'*z']$
    \item $\plr{\exists y.~{y \leq}0 \land {x \leq}[y+z]}
      \leftrightarrow x \leq z$
    \item $\plr{\exists z.~{x \leq}[y+z] \land {z \leq}0}
      \leftrightarrow x \leq y$
    \item $\plr{\exists wx.~{wx \leq}[w+x] \land {z \leq}[wx+y]}$
      $\leftrightarrow$

      $\plr{\exists xy.~{z \leq}[w+xy] \land {xy \leq}[x+y]}$
    \item $\exists x.~{x \leq}0 \land \forall x'.~{x' \leq}0 \to x' \leq x$
    \item $\vdots$
    %\item $x \leq x' \to y \leq y' \to x + y \leq x' + y'$
    %\item $x \leq x' \to y \leq y' \to x * y \leq x' * y'$
    %\item $0 + x = x \land x + 0 = x$
    %\item $(x + y) + z = x + (y + z)$
    %\item $x + y = y + x$
    %\item $1 * x = x \land x * 1 = x$
    %\item $(x * y) * z = x * (y * z)$
    %\item $0 * x = 0 \land x * 0 = 0$
    %\item $(x + y) * z = x * z + y * z$
    %\item $x * (y + z) = x * y + x * z$
  \end{itemize}
\end{definition}

@@

>>>>

@@

Vectors of a given length form a \emph{module} over the posemiring $\Ann$,
analogously to how vectors over a field form a vector space.

\begin{definition}
  A \emph{(left) module over a posemiring}, given a posemiring $\Ann$, is a
  partially ordered commutative monoid $(M, 0_M, +_M)$ with, for each
  $r \in \Ann$, a pomonoid morphism $r \cdot \plr{-} : M \to M$, such that the
  collection of these respects the posemiring structure on $r$.
  Specifically:
  \begin{itemize}
    \item $r \leq r' \to u \leq u' \to r \cdot u \leq r' \cdot u'$
    \item $r \cdot 0_M = 0_M$
    \item $r \cdot \plr{u +_M v} = r \cdot u +_M r \cdot v$
    \item $0 \cdot u = 0_M$
    \item $\plr{r + s} \cdot u = r \cdot u +_M s \cdot u$
    \item $1 \cdot u = u$
    \item $\plr{r * s} \cdot u = r \cdot \plr{s \cdot u}$
  \end{itemize}
\end{definition}

We care to define modules so as to define \emph{module morphisms}, also known
as \emph{linear maps}, which we use extensively when relating two contexts (as
we do, for example, in simultaneous substitution).
For the sake of mechanisation, we choose to define module morphisms
\emph{relationally} rather than \emph{functionally}, giving a somewhat
unfamiliar-looking definition that is equivalent to the usual functional
definition.
The main advantage of this relational approach is that proofs of relatedness
for typical linear maps compose and decompose via data constructors and
pattern matching.
% I'm not sure this is true:
%An auxiliary advantage is that with relations rather than functions, we can
%much more often take advantage of judgemental injectivity, thus making
%unification-based solving of implicits more effective.
%For example, if \AgdaBound{R} is a free variable of relation type, then
%\AgdaInductiveConstructor{refl} serves as a proof of
%\ExecuteMetaData[\Snippetstex]{Rxy-R}{}, solving the underscores as
%\AgdaBound{x} and \AgdaBound{y}, respectively.
